<<<<<<< HEAD
---
title: "R Notebook"
output: html_notebook
---

```{r Load Library, warning=FALSE,message=FALSE}

# Clear Environment
rm(list=ls())

# Load Libraries
require(terra)
require(tidyverse)
library(geoR)
require(spatstat)
require(sf)
require(tidyterra)
require(cowplot)
require(plotGLM)

# Packages used, but not loaded due to function conflicts
# require(pgirmess)
# require(ncf)

```


# Background

Researchers set up a square plot with a dimension of 130 m on each side. Within the plot they established blue vane traps in a 5 m grid pattern to sample bee abundance. Every day for 10 days the researchers emptied the blue vane traps and recorded the number of bees collected. After the 10-day sampling period, they calculated the mean daily number of bees captured in the trap each day.

# Challenge 1 (2 points)

Import the blueVanes.csv file from GitHub. Plot the mean bee abundance as a function of the x and y coordinates. Upon initial inspection, does there appear to be spatial dependence in the bee response data?

```{r}
# Import Blue Vane Trap Data
blueVane = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week6/blueVanes.csv')

# Plot Mean Bee Abundance
C1plot =  ggplot(blueVane, aes(x=x, y=y, color = meanBee))+
  geom_point(size = 5)+
  theme_bw()+
  theme(panel.grid = element_blank())+
  scale_color_gradientn(colors = c('blue','orange','red'),name = 'Abundance')+
  labs(title = "Mean Bee Abundance", y = "Y Coordinates", x = "X Coordinates")+
  theme(plot.title = element_text(face = 'bold', size = 14, hjust = 0.5),
        axis.title = element_text(face = 'bold', size = 12),
        legend.title = element_text(face = 'bold'))

C1plot

```

**Challenge 1 Answer:**

There are a few areas were we have some clustering of low values (lower-left and upper right quadrants) and a few areas where there is some clustering of high values (lower-right and middle-left).  Generally I would say there is some evidence there is spatial dependence in mean bee abundance across the sample area


# Challenge 2 (3 points)

Empirically test for evidence of spatial dependence in bee abundance by first building a corelogram, then building a semi-variogram. Be sure to provide an envelope for assessing whether the values in each figure are statistically different from what would be expected if the bee values were randomly distributed. Describe the patterns you see, and your interpretation with regards to the spatial dependence in mean bee abundance. Be sure to identify the maximum distance at which you should interpret your figures (2/3 of the maximum distance between points).


```{r Challenge 2, warning=F,message=F}

# Create Distance Matrix to determine max analysis distance
blueVaneCoords = cbind(blueVane$x,blueVane$y)
colnames(blueVaneCoords) = c('x','y')
blueVaneDistMatrix = as.matrix(dist(blueVaneCoords))

# # Limit Max Distance
maxDist = (2/3)*max(blueVaneDistMatrix)


# Correlogram ncf
blueVaneNcf = ncf::correlog(x = blueVane$x, y=blueVane$y, z=blueVane$meanBee, increment=5, resamp=100)

# Create Temporary Data frame to store Correlogram outputs
tmp = cbind(blueVaneNcf$n, blueVaneNcf$mean.of.class, blueVaneNcf$correlation,blueVaneNcf$p) %>% 
  data.frame() %>%
  set_names(c('n', 'dist', 'corr', 'p')) %>% 
  mutate(significant = ifelse(p < 0.05, 'yes', 'no'))

# Plot the Correlogram
ggplot(tmp, aes(x=dist, y=corr))+
  geom_line()+
  geom_point(aes(color=significant, size=n))+
  scale_color_manual(values=c('black', 'red'), name = 'Significant?')+
  scale_size(name = 'Sample Size')+
  theme_bw()+
  theme(panel.grid=element_blank())+
  geom_hline(yintercept=0, linetype='dashed')+
  geom_vline(xintercept=maxDist, color='red')+
  scale_x_continuous(breaks=seq(0, maxDist+5, 5))+
  labs(title = 'Challenge 2 Correlogram', y = 'Correlation', x = 'Distance')+
  theme(plot.title = element_text(size = 14, hjust = 0.5, face = 'bold'),
        axis.title = element_text(size = 12, face = 'bold'),
        legend.title = element_text(face = 'bold'))+
  coord_cartesian(xlim = c(0,maxDist+5), ylim = c((min(tmp$corr)-0.05),0.3))
  
# Semi-Variogram

# Use this function to compare model based variograms
geoBlueVane = as.geodata(blueVane)

# Semi-Variogram
empGeor = variog(geoBlueVane,max.dist = maxDist, breaks = seq(0,maxDist,by = 5))
plot(empGeor, main = 'Challenge 2 Semi-Variogram')

# Visually Inspect Semi-variogram from initialization values
sillSV = 13
rangeSV = 30

# Test for Anisotropy
emp4Geor = variog4(geoBlueVane, max.dist=maxDist, direction = c(0, pi/4, pi/2, 3*pi/4))
plot(emp4Geor)
title(main = 'Challenge 2 Test of Anisotropy')

# Fit Model to Semi-variogram with exponential and shperical models
mlExp = likfit(geoBlueVane, cov.model='exp', ini=c(sillSV, rangeSV))
mlSph = likfit(geoBlueVane, cov.model='sph', ini=c(sillSV, rangeSV))

# Compare Models - extract model with lowest AIC
resultsAIC = AIC(mlExp, mlSph)

plot(empGeor, main = 'Challenge 2 Competing Models for Semi-Variogram')
lines(mlExp,col = 'blue')
lines(mlSph,col = 'red')

# Monte Carlo Simulation to Test against Random
empEnv = variog.mc.env(geoBlueVane, obj.var=empGeor, nsim=99)
plot(empGeor, envelope=empEnv, main = 'Challenge 2 Monte Carlo Simulation of Significance')
lines(mlSph, col='blue')

```

**Challenge 2 Answer:**

Based on the empirical tests, there is evidence for spatial dependence in the data set. First, the correlegram indicates there is significant correlation between sample locations that are within 25 meters of each other. 

The top performing model, based on AIC, of the trends in the semi-variogram was the spherical model with an estimated nugget of 9.2, and a range of 41.87. Monte Carlo simulations corroborate the inference from the correlogram, suggesting that there is similarities in the variance of mean bee abundance measurements between points within roughly 25 meters.

I tested for anisotropy, and concluded that trends in spatial dependence were similar in all directions.

Lastly, interpretations were limited to 90 meters, as this value represents 2/3 of the maximum distance between points.

Generally, if we fail to account for the spatial dependence in the data, our estimates of mean bee abundance may be biased, and/or the risk of type 1 errors increases.


# Challenge 3 (4 points)

Fit a simple linear model (assuming a normally distributed error term) that models mean bee abundance as a function of an intercept only (no other covariates in the model). Compare the model-estimated intercept to the bee abundance values averaged across all sites. What do you notice?  

Empirically test for evidence of spatial autocorrelation IN THE MODEL RESIDUALS by first building a correlogram, then building a semi-variogram. Be sure to provide an envelope for assessing whether the values in each figure are statistically different from what would be expected if the bee values were randomly distributed in space. Compare these figures to those built for the raw data. What similarities or differences to you notice?

```{r Challenge 3, warning=FALSE, message=FALSE}
# Linear Model
linearBlueVane = glm(blueVane$meanBee ~ 1)
linearBlueVane$coefficients
mean(blueVane$meanBee)

# Residuals
blueVaneResiduals = as.data.frame(cbind(blueVaneCoords,linearBlueVane$residuals))
colnames(blueVaneResiduals) = c('x','y','resid')

# Simple Plot
C3plot = ggplot(blueVaneResiduals, aes(x=x, y=y, color = resid))+
  geom_point(size = 5)+
  theme_bw()+
  theme(panel.grid = element_blank())+
  scale_color_gradientn(colors = c('blue','orange','red'),name = 'Error')+
  labs(title = "Residuals in Linear Model of Mean Bee Abundance", y = "Y Coordinates", x = "X Coordinates")+
  theme(plot.title = element_text(face = 'bold', size = 14, hjust = 0.5),
        axis.title = element_text(face = 'bold', size = 12),
        legend.title = element_text(face = 'bold'))

C3plot

# Correlogram
blueVaneResidNcf = ncf::correlog(x = blueVaneResiduals$x, y=blueVaneResiduals$y, z=blueVaneResiduals$resid, increment=5, resamp=100)

tmp = cbind(blueVaneResidNcf$n, blueVaneResidNcf$mean.of.class, blueVaneResidNcf$correlation,
            blueVaneResidNcf$p) %>% 
  data.frame() %>% set_names(c('n', 'dist', 'corr', 'p')) %>% 
  mutate(significant = ifelse(p < 0.05, 'yes', 'no'))

# Plot the Correlogram
ggplot(tmp, aes(x=dist, y=corr))+
  geom_line()+
  geom_point(aes(color=significant, size=n))+
  scale_color_manual(values=c('black', 'red'), name = 'Significant?')+
  scale_size(name = 'Sample Size')+
  theme_bw()+
  theme(panel.grid=element_blank())+
  geom_hline(yintercept=0, linetype='dashed')+
  geom_vline(xintercept=maxDist, color='red')+
  scale_x_continuous(breaks=seq(0, maxDist+5, 5))+
  labs(title = 'Challenge 3 Correlogram', y = 'Correlation', x = 'Distance')+
  theme(plot.title = element_text(size = 14, hjust = 0.5, face = 'bold'),
        axis.title = element_text(size = 12, face = 'bold'),
        legend.title = element_text(face = 'bold'),
        plot.margin = unit(c(0,0,0,0),'cm'))+
  coord_cartesian(xlim = c(0,maxDist+5), ylim = c((min(tmp$corr)-0.05),0.3))


# Semi-Variogram

# Use this function to compare model based variograms
geoBlueVaneResid = as.geodata(blueVaneResiduals)

# Semi-Variogram
empGeor = variog(geoBlueVaneResid, max.dist = maxDist, breaks = seq(0,maxDist,by = 5))
plot(empGeor, main = 'Challenge 3 Semi-Variogram')

# Visually Inspect Semi-variogram from initialization values
sillSV = 13
rangeSV = 30

# Test for Anisotropy
emp4Geor = variog4(geoBlueVaneResid, max.dist=maxDist, direction = c(0, pi/4, pi/2, 3*pi/4))
plot(emp4Geor)
title(main = 'Challenge 3 Test of Anisotropy')

# Fit Model to Semi-variogram with exponential and shperical models
mlExp = likfit(geoBlueVaneResid, cov.model='exp', ini=c(sillSV, rangeSV))
mlSph = likfit(geoBlueVaneResid, cov.model='sph', ini=c(sillSV, rangeSV))

# Compare Models - extract model with lowest AIC
resultsAIC = AIC(mlExp, mlSph)

plot(empGeor, main = 'Challenge 3 Competing Models for Semi-Variogram')
lines(mlExp,col = 'blue')
lines(mlSph,col = 'red')

# Monte Carlo Simulation to Test against Random
empEnv = variog.mc.env(geoBlueVaneResid, obj.var=empGeor, nsim=99)
plot(empGeor, envelope=empEnv, main = 'Challenge 3 Monte Carlo Simulation of Significance')
lines(mlSph, col='blue')

```

**Challenge 3 Answer:**
The coefficient estimate for the intercepts is the average mean bee counts across all sites.

The correlogram and semi-variogram based on the residuals is very similar to the correlogram based on the raw data.

Together the empirical tests suggest there is spatial dependence in the residual errors of the linear model we used to estimate mean bee abundance. This represents a violated assumption in our linear regression model, which may results in biased parameter estimates or inflated type 1 error risk.


# Challenge 4 (4 points)

Conventional wisdom indicates that the distribution of floral resources may substantially influence bee abundance. In addition, I am  completely making up the following fact: the average bee has a perceptual range of about 15 m. Below I have provided code for importing a raster that shows the number of flowers counted in 1 m grid cells across the entire study area. Use the focal() function from the previous lab to generate a new raster in which each cell value represents the mean number of flowers in cells within a 15 m radius. Plot the distribution of the blue vane traps over the 15 m focal raster. No text required here, just show me the plot.

```{r Challenge 4, warning=FALSE, message=FALSE}
# Import Floral Resources Layer
floral = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week6/floralResources.tif')
# Verify 1m x 1m Resolution
res(floral)
# Quick Visual of Floral Resources Layer
plot(floral, main = 'Floral Resources')

# Create spatvector for plotting
blueVaneSpatVect = vect(blueVane, geom = c('x','y'))

# Focal Statistics to summarize floral resources within 15m radius
focalfloral = focal(floral,w=15, fun = 'mean')

# Plot of resources relative to blue vane trap locations
plot(focalfloral, main = 'Mean Floral Resources within 15M')
plot(blueVaneSpatVect, alpha = 0.65, pch = 3, extend = FALSE,
     add = T)


```



# Challenge 5 (4 points)

Extract the focal flower value from the raster you just created at each point using the extract() function. You now have an estimate for the average number of flowers per meter around each blue vane trap. Fit a second linear model, but this time include average floral abundance within 15 m as a covariate. Export the residuals from this fitted model, and again, examine a corelogram and variogram of the residuals. As always, be sure to plot random envelopes. Now what do you notice?

```{r}
# Extract Raster Value, append to coordinates of blue vane trap sites
focalExtract = terra::extract(focalfloral, blueVaneCoords )

# Combine trap coordinates with mean resource value
blueVaneFocalExtract = as.data.frame(cbind(blueVane,focalExtract))

# glm with floral resources covariate
frGLM = glm(blueVaneFocalExtract$meanBee ~ blueVaneFocalExtract$focal_mean)

# Extract residuals from glm
focalResid = frGLM$residuals

# Combine residuals, coordinates and mean floral resources in a dataframe
blueVaneFocalResiduals = cbind(blueVaneFocalExtract[,1:2], focalResid)

# Create Correlogram
blueVaneFocalResidNcf = ncf::correlog(x = blueVaneFocalResiduals$x, y=blueVaneFocalResiduals$y, z=blueVaneFocalResiduals$focalResid, increment=5, resamp=100)

tmp = cbind(blueVaneFocalResidNcf$n, blueVaneFocalResidNcf$mean.of.class, blueVaneFocalResidNcf$correlation,
            blueVaneFocalResidNcf$p) %>% 
  data.frame() %>% set_names(c('n', 'dist', 'corr', 'p')) %>% 
  mutate(significant = ifelse(p < 0.05, 'yes', 'no'))

ggplot(tmp, aes(x=dist, y=corr))+
  geom_line()+
  geom_point(aes(color=significant, size=n))+
  scale_color_manual(values=c('black', 'red'))+
  theme_bw()+
  theme(panel.grid=element_blank())+
  geom_hline(yintercept=0, linetype='dashed')+
  geom_vline(xintercept=maxDist, color='red')+
  scale_x_continuous(breaks=seq(0, 130, 5))+
  labs(title = 'Challenge 5 Correlogram', y = 'Correlation', x = 'Distance')+
  theme(plot.title = element_text(size = 14, hjust = 0.5, face = 'bold'),
        axis.title = element_text(size = 12, face = 'bold'),
        legend.title = element_text(face = 'bold'),
        plot.margin = unit(c(0,0,0,0),'cm'))+
  coord_cartesian(xlim = c(0,maxDist+5), ylim = c((min(tmp$corr)-0.05),0.3))


# Semi-Variogram

# Use this function to compare model based variograms
geoBlueVaneFocalResid = as.geodata(blueVaneFocalResiduals)

# Semi-Variogram
empGeor = variog(geoBlueVaneFocalResid, max.dist = maxDist, breaks = seq(0,maxDist,by = 5))
plot(empGeor, main = 'Challenge 5 Semi-Variogram')

# Visually Inspect Semi-variogram from initialization values
sillSV = 10
rangeSV = 15

# Test for Anisotropy
emp4Geor = variog4(geoBlueVaneFocalResid, max.dist=maxDist, direction = c(0, pi/4, pi/2, 3*pi/4))
plot(emp4Geor)
title(main = 'Challenge 5 Test of Anisotropy')

# Fit Model to Semi-variogram with exponential and shperical models
mlExp = likfit(geoBlueVaneFocalResid, cov.model='exp', ini=c(sillSV, rangeSV))
mlSph = likfit(geoBlueVaneFocalResid, cov.model='sph', ini=c(sillSV, rangeSV))

# Compare Models - extract model with lowest AIC
resultsAIC = AIC(mlExp, mlSph)

plot(empGeor, main = 'Challenge 5 Competing Models for Semi-Variogram')
lines(mlExp,col = 'blue')
lines(mlSph,col = 'red')

# Monte Carlo Simulation to Test against Random
empEnv = variog.mc.env(geoBlueVaneFocalResid, obj.var=empGeor, nsim=99)
plot(empGeor, envelope=empEnv, main = 'Challenge 5 Monte Carlo Simulation of Significance')
lines(mlSph, col='blue')

```

**Challenge 5 Answer:**

After including the mean floral resources within 15 meters of the sample site, we have effectively eliminated the issue of spatial dependence. Correlation and variation between the measure of the residual error in mean bee abundance is consistent across all measured distances. 

# Challenge 6 (3 points)

What is your assessment regarding the distribution of bee abundance in this study area? Does it exhibit spatial dependence, and if so, what is driving that pattern?

**Challenge 6 Answer:**

The distribution of mean bee abundance in this study area does exhibit spatial dependence. The primary driver of the pattern was floral resources. Once we addressed the issue of spatial dependence, by including mean floral resources within 15 meters of each blue vane trap in the model, our estimates of the effect size and error are more reliable. 

Overall, we see a positive relationship between mean bee abundance and floral resources. As floral resources increase within the 15 meter radius of the blue vane trap, mean bee abundance increases. The intercept of the model is 4.25, suggesting that mean bee abundance across the study site is 4.25. This estimate is roughly one third of the estimate produced by the original intercept only model. 
